---
title: "Class10MiniProject"
format: html
author: Benjamin Gersh
editor: visual
---

```{r}
candy_file <- "candy-data.csv"
candy = read.csv(candy_file, row.names=1)
head(candy)
```

>Q1.How many different types of candy are in this dataset?

```{r}
nrow(candy)
```
85 different types of candy are in this dataset. 

>Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```
There are 38 fruity candy types in the dataset. 

>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Air Heads", ]$winpercent
```
The winpercent value for Air Heads is 52.34.

>Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```
76.77

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

```{r}
library("skimr")
skim(candy)
```

>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes, any column that uses percent seems to be on a continuous scale whereas non percent is 0 and 1. 


>Q7. What do you think a zero and one represent for the candy$chocolate column?

A zero represents that it fits in the category or not, like there is chocolate or there is no chocolate in the category.

>Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```


```{r}
library(ggplot2)

ggplot(candy) +
aes(winpercent) +
 geom_histogram(bins=10, col="red", fill="orange")
```

>Q9. Is the distribution of winpercent values symmetrical?

No, the distribution of winpercent values is not symmetrical.

>Q10.Is the center of the distribution above or below 50%?

The center of the distribution i below 50%. 

>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
chocolate.inds<-as.logical(candy$chocolate)
chocolate.wins<-candy[chocolate.inds,]$winpercent
mean(chocolate.wins)
```
```{r}
fruity.inds<-as.logical(candy$fruity)
fruity.wins<-candy[fruity.inds,]$winpercent
mean(fruity.wins)
```
On average, chocolate candy is higher ranked than fruity candy.

>Q12. Is this difference statistically significant?

```{r}
t.test(chocolate.wins, fruity.wins)
```
Yes, the difference is statistically significant due to the small p-value.

>Q13. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent),], n=5)
```


>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
tail(candy[order(candy$winpercent),], n=5)
```

>Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?
```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

>Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate candy is Sixlets.
>Q18. What is the best ranked fruity candy?

Starburst is the best ranked fruity candy.

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 20)
```
>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

The highest ranked candy in terms of winpercent for the least money is Reese's Miniatures. 

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )

```

The top 5 most expensive candy types are Nik L Nip, Ring Pop, Hershey's Krackel, and Hershey's Milk Chocolate. Out of these 5, the least popular is Nik L Nip. 

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity

>Q23. Similarly, what two variables are most positively correlated?

Chocolate and bar

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```
```{r}
plot(pca$x[,1:2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)
p

```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 34)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
library(plotly)
```

```{r}
ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard, and pluribus are the original variables that are picked up strongly by PC1 in the positive direction. These make sense to me because these are typically associated with one another and not associated with the PC1 in the negative direction. For example, fruity and chocolate are typically not associated with each other, which is shown by the PC1 data. 